---
title: "Network analysis"
output: html_notebook
---

```{r include=FALSE}
install.packages("igraph")
library(igraph) 
```

## Networks in igraph
[Source](http://kateto.net/ruworkshop#1_introduction_to_r)

### Create networks

Generate an undirected graph with three edges. The numbers are interpreted as vertex IDs, so the edges are `1–>2`, `2–>3`, `3–>1`.

```{r}
g1 <- graph( edges=c(1,2, 2,3, 3, 1), n=3, directed=F ) 
plot(g1)
```
```{r}
class(g1)
```

```{r}
g1
```

Now with 10 vertices, and directed by default:

```{r}
g2 <- graph( edges=c(1,2, 2,3, 3, 1), n=10 )
plot(g2)
```

```{r}
g2
```

Named vertices

```{r}
g3 <- graph( c("John", "Jim", "Jim", "Jill", "Jill", "John")) 
plot(g3)
```

```{r}
g3
```

In named graphs we can specify isolates by providing a list of their names.

```{r}
g4 <- graph( c("John", "Jim", "Jim", "Jack", "Jim", "Jack", "John", "John"), 
             isolates=c("Jesse", "Janis", "Jennifer", "Justin") )  

plot(g4, edge.arrow.size=.5, vertex.color="gold", vertex.size=15, 
     vertex.frame.color="gray", vertex.label.color="black", 
     vertex.label.cex=0.8, vertex.label.dist=2, edge.curved=0.2) 
```

Small graphs can also be generated with a description of this kind: `-` for undirected tie, `+-` or `-+` for directed ties pointing left & right, `++` for a symmetric tie, and `:` for sets of vertices.

```{r}
plot(graph_from_literal(a---b, b---c))
```

```{r}
plot(graph_from_literal(a--+b, b+--c))
```

```{r}
plot(graph_from_literal(a+-+b, b+-+c)) 
```

```{r}
plot(graph_from_literal(a1:a2:a3---b1:b2:b3))
```

```{r}
gl <- graph_from_literal(a-b-c-d-e-f, a-g-h-b, h-e:f:i, j)
plot(gl)
```

### Edge, vertex, and network attributes

Access vertices and edges

```{r}
plot(g4)
E(g4)
```

```{r}
V(g4)
```

Examine the network matrix
```{r}
g4[]
```

```{r}
g4[1,]
```

Add attributes to the network, vertices, or edges

```{r}
V(g4)$name
```

```{r}
V(g4)$gender <- c("male", "male", "male", "male", "female", "female", "male")
E(g4)$type <- "email" 
E(g4)$weight <- 10

edge_attr(g4)
```

```{r}
vertex_attr(g4)
```

```{r}
graph_attr(g4)
```

Another way to set attributes (you can similarly use `set_edge_attr()`, `set_vertex_attr()`, etc.):
```{r}
g4 <- set_graph_attr(g4, "name", "Email Network")
g4 <- set_graph_attr(g4, "something", "A thing")

graph_attr_names(g4)
```

```{r}
graph_attr(g4, "name")
```

```{r}
graph_attr(g4)
```

```{r}
g4 <- delete_graph_attr(g4, "something")
graph_attr(g4)
```

```{r}
plot(g4, edge.arrow.size=.5, vertex.label.color="black", vertex.label.dist=1.5,
     vertex.color=c( "pink", "skyblue")[1+(V(g4)$gender=="male")] ) 
```

The graph `g4` has two edges going from `Jim` to `Jack`, and a loop from `John` to himself. We can simplify our graph to remove loops & multiple edges between the same nodes. Use `edge.attr.comb` to indicate how edge attributes are to be combined - possible options include `sum`, `mean`, `prod` (product), `min`, `max`, `first`/`last` (selects the first/last edge’s attribute). Option `ignore` says the attribute should be disregarded and dropped.

```{r}
g4s <- simplify( g4, remove.multiple = T, remove.loops = F, 
                 edge.attr.comb=c(weight="sum", type="ignore") )
plot(g4s, vertex.label.dist=1.5)
```

```{r}
g4s
```

The description of an igraph object starts with up to four letters:

1. `D` or `U`, for a directed or undirected graph
2. `N` for a named graph (where nodes have a `name` attribute)
3. `W` for a weighted graph (where edges have a `weight` attribute)
4. `B` for a bipartite (two-mode) graph (where nodes have a `type` attribute)

The two numbers that follow (`7 3`) refer to the number of nodes and edges in the graph. The description also lists node & edge attributes, for example:

* (`g/c`) - graph-level character attribute
* (`v/c`) - vertex-level character attribute
* (`e/n`) - edge-level numeric attribute

### Specific graph and graph models 

**Empty graph**
```{r}
eg <- make_empty_graph(40)
plot(eg, vertex.size=10, vertex.label=NA)
```

**Full graph**
```{r}
fg <- make_full_graph(40)
plot(fg, vertex.size=10, vertex.label=NA)
```

**Simple star graph**
```{r}
st <- make_star(40)
plot(st, vertex.size=10, vertex.label=NA) 
```

**Tree graph**
```{r}
tr <- make_tree(40, children = 3, mode = "undirected")
plot(tr, vertex.size=10, vertex.label=NA) 
```

**Ring Graph**
```{r}
rn <- make_ring(40)
plot(rn, vertex.size=10, vertex.label=NA)
```

**Erdos-Renyi random graph model**
`n` is number of nodes, `m` is the number of edges
```{r}
er <- sample_gnm(n=100, m=40) 
plot(er, vertex.size=6, vertex.label=NA) 
```

**Watts-Strogatz small-world model**
Creates a lattice (with `dim` dimensions and `size` nodes across dimension) and rewires edges randomly with probability `p`. The neighborhood in which edges are connected is `nei`. You can allow loops and multiple edges.
```{r}
sw <- sample_smallworld(dim=2, size=10, nei=1, p=0.1)
plot(sw, vertex.size=6, vertex.label=NA, layout=layout_in_circle)
```

**Barabasi-Albert preferential attachment model for scale-free graphs**
`n` is number of nodes, `power` is the power of attachment (1 is linear); `m` is the number of edges added on each time step
```{r}
ba <-  sample_pa(n=100, power=1, m=1,  directed=F)
plot(ba, vertex.size=6, vertex.label=NA)
```

**Notable historical graphs**
```{r}
zach <- graph("Zachary") # the Zachary carate club
plot(zach, vertex.size=10, vertex.label=NA)
```

**Rewiring a graph**
`each_edge()` is a rewiring method that changes the edge endpoints uniformly randomly with a probability `prob`.
```{r}
rn.rewired <- rewire(rn, each_edge(prob=0.1))
plot(rn.rewired, vertex.size=10, vertex.label=NA)
```

Rewire to connect vertices to other vertices at a certain distance.
```{r}
rn.neigh = connect.neighborhood(rn, 5)
plot(rn.neigh, vertex.size=8, vertex.label=NA) 
```

Combine graphs (disjoint union, assuming separate vertex sets) `%du%`
```{r}
plot(rn, vertex.size=10, vertex.label=NA) 
```

```{r}
plot(tr, vertex.size=10, vertex.label=NA) 
```

```{r}
plot(rn %du% tr, vertex.size=10, vertex.label=NA) 
```

## Reading network data
